---
title: "Birth Weights"
author: "Léon Yuan"
format: pdf
editor: visual
---

# Explore Data Analysis

## How many Black in this data?

It is quite balanced variable. There are 662 black mother and 453 non-black mother.

```{r}
table(df$black)
```

## Education year distribution by black mother and non-black

From this density curve by black or not, we can tell that both groups have around 12 years of education to be mode/peak. However, black mothers density curve has sharper multiple modes such as 9, 10, 11, 13, 14, 15. Black mothers have much more education of 12 years than non-black mothers do while non-black mothers have more of education years more than 15 than black mothers.

```{r}
#| message: false
library(tidyverse)
cols <- c("#F76D5E", "#FFFFBF")
# Basic density plot in ggplot2
ggplot(df, aes(x = educ, fill = black)) +
  geom_density(alpha = 0.7) + 
  scale_fill_manual(values = cols) 
```

## How many smoke/non-smoke in black/non-black mother?

In general, there are more black smoker mother than non-black mothers. Within the black group and non-black group, there is higher rate of smokers in black than non-black.

```{r}
#| message: false
library(scales)
df |>
  count(black, smoke) |>
  group_by(black) |>
  mutate(prop = n / sum(n)) |>
  ggplot(aes(x = black, fill = smoke, y = n,
             label = percent(prop, accuracy = 0.1))) +
  geom_bar(position="dodge", stat="identity") +
  geom_text(position = position_dodge(width = .9),# move to center of bars
              vjust = -0.5,    # nudge above top of bar
              size = 4) +
  ggtitle("Smoke/Non-smoke by black or not")
```

## Gestate distribution by black

This shows gestate is very skewed distributed, this violates the linear regression assumptions of Multivariate normality.

```{r}
cols <- c("#F76D5E", "#FFFFBF")
# Basic density plot in ggplot2
ggplot(df, aes(x = gestate, fill = black)) +
  geom_density(alpha = 0.7) + 
  scale_fill_manual(values = cols)
```

# Fit a ordinary linear regression model

The ordinary linear regression performs very poor on this data set because the adjusted R squared is only 52.61%.

```{r}
fit <- lm(grams ~., data = df)
summary(fit)
```

The residuals closely form a normal distribution, however, residuals are very heteroscedasticity, as the fitted values increases, the variance of residuals increases as well radically.

```{r}
plot(fit)
extractAIC(fit, k=log(nrow(df)))
```

Take a close look at the distribution of grams:

```{r}
plot(density(df$grams))
```

```{r}
sum(df$grams < 2000)
```

# Take a log transformation on the grams variable then OLS

```{r}
fit2 <- lm(log(grams) ~., data = df)
summary(fit2)
```

```{r}
plot(fit2)
extractAIC(fit2, k=log(nrow(df)))
```

# Take a log on y and add interaction between black and smoke

```{r}
fit3 <- lm(log(grams) ~ black + black:smoke + smoke + educ + gestate, data = df)
summary(fit3)
```

```{r}
plot(fit3)
```

# Fit a model with all possible interaction

```{r}
fit4 <- lm(log(grams) ~ (black + educ + smoke + gestate)^2, df)
summary(fit4)
```

```{r}
plot(fit4)
```

# Use BIC to backforward select the best model

```{r}
model_bic <- step(fit4, direction = "backward", k=log(nrow(df)))
```

```{r}
fit5 <- lm(log(grams) ~ black + educ + smoke + gestate + black:educ + black:smoke + black:gestate + educ:gestate + smoke:gestate, df)
summary(fit5)
```

```{r}
plot(fit5)
```

# Make box-cox transformation on y

```{r}
plot(density(log(df$grams)))
```

# Try random forest

```{r}
#| message: false
library(randomForest)
library(datasets)
library(caret)
library(pROC)
library(glmnet)
```

```{r}
set.seed(222)
ind <- sample(2, nrow(df), replace = TRUE, prob = c(0.8, 0.2))
train <- df[ind==1,]
test <- df[ind==2,]
dim(train)
dim(test)
```

```{r}
rf <- randomForest(x = train[-5],
                   y = log(train$grams),
                   xtest = test[-5],
                   ytest = log(test$grams),
                   ntree = 500,
                   mtry = 2,
                   proximity = TRUE)
print(rf)
```

```{r}
varImpPlot(rf,
           sort = T,
           n.var = 4,
           main = "Top 4 - Variable Importance")
importance(rf)
```

```{r}
fit6 <- lm(log(grams) ~ black + educ + smoke + gestate + black:educ + black:smoke + black:gestate + educ:gestate + smoke:gestate, train)
summary(fit6)
pred <- predict(fit6, newdata = test[-5])
mean((pred-log(test$grams))^2)
```

```{r}
plot(density(log(df$gram)))
boxcox(grams^2 ~ black + educ + smoke + log(gestate), data = df)
```

```{r}
set.seed(123) 
a = 0.3
f = "gaussian"
# Encode matrix into dummy variable forms for all categorical variables
x.train <- model.matrix(log(grams) ~ (black + educ + smoke + log(gestate))^2,
                        train)[,-1]
# Use Elastic net with logistics regression on this train dataset with alpha = 0.5
cv.elastic <- cv.glmnet(x = x.train, y = log(train$grams), 
                      alpha = a, family = f)
# Build a final model with the best lambda selected by cross validation measured by Binomial Deviance
best_elastic <- glmnet(x.train, y = log(train$grams), alpha = a, family = f,
                lambda = cv.elastic$lambda.1se)
coef(best_elastic)
plot(cv.elastic, main = "Cross-Validation to select shrinkega lamda")
```

```{r}
x.test <- model.matrix(log(grams) ~ (black + log(educ) + smoke + log(gestate))^2,
                        test)[,-1]
prob_elastic <- predict(best_elastic, newx = x.test, type = "response")
mean((prob_elastic-log(test$grams))^2)
```

```{r}
fit7 <- glm(grams ~ (black + educ + smoke + log(gestate))^2,
            data = df,
            family = inverse.gaussian(link = "inverse"))
summary(fit7)
plot(x=predict(fit7, type = "link"), y=residuals(fit7, "pearson"))
```

```{r}
fit8 <- glm(log(grams) ~ (black + educ + smoke + log(gestate))^2,
            data = train,
            family = "gaussian")
for (i in 1:20)
{
  w <- 1/predict(fit8, type = "response")^2
  fit8 <- glm(log(grams) ~ (black + educ + smoke + log(gestate))^2,
                      data = train,
                      family = "gaussian",
                      weights = w)
}
plot(x=predict(fit8, type = "link"), y=residuals(fit8, "pearson"))
pred_fit8 <- predict(fit8, newdata = test, type = "response")
mean((pred_fit8-log(test$grams))^2)
View(cbind(test, exp(pred_fit8)))
```

```{r}
fit9 <- rlm(log(grams) ~ (black + educ + smoke + log(gestate))^2,
            data = train,
            maxit = 100, psi = psi.bisquare)
for (i in 1:20)
{
  w <- 1/predict(fit9, type = "response")^2
  fit9 <- rlm(log(grams) ~ (black + educ + smoke + log(gestate))^2,
                      data = train,
                      maxit = 100, 
                      psi = psi.bisquare,
                      weights = w)
}
plot(x=predict(fit9, type = "response"), y=residuals(fit9, "pearson"))
pred_fit9 <- predict(fit9, newdata = test, type = "response")
mean((pred_fit9-log(test$grams))^2)
View(cbind(test, exp(pred_fit9)))
```

